A Geometry for Network Visualization in \MakeLowercase{\texttt{ggplot2}}

### 2015-03-21 CC/writeup.Rnw
edge id

@@

This network is shown in figure~\ref{fig.cap:madmen_ex}. \emph{Notes: add names, color for gender} In the plot, we can see there is one central character who has many more relationships than any other character. This vertex represents the main character of the show, Don Draper.
\par There are many kinds of networks, and networks are extensively studied across many disciplines.  Many sociologists study social networks, and many biologists study protein networks. As different as these and the many other disciplines that study networks are, they all need the ability to quickly and effectively visualize networks.  We found the current tools to be lacking in this ability, so we chose to fill this gap by adding network plotting capabilities to the popular and widely used \texttt{R} package \texttt{ggplot2}

@@

\par The \texttt{GGally} package by \citet{ggally} contains a very useful function written by Fran\c{c}ois Briatte and Moritz Marbach called \texttt{ggnet} that does allow for fairly straightforward customization of these three necessary graph attributes. \hh{XXX drive home the message that vertices, edges and layout are important, and repeat at this point} \sctq{Not quite sure what to say here. What follows is my guesstimate.} \sct{Coloring the vertices or edges in a graph is a quick and easy way to visualize grouping and can help with pattern or cluster detection. The vertices in a network and the edges between them compose the structure of a network, and being able to discover patterns among them visually is a key part of network analysis. Viewing multiple layouts of the same network can also help reveal patterns or clusters that would not be discovered when only viewing one layout or analyzing only an adjacency matrix.} The \texttt{ggnet} function, however, requires the graph input to be a \texttt{network} object according to the \texttt{network} package.  Our goal is to build off of the \texttt{ggnet} function and present an intuitive \texttt{geom} for network visualization within the \texttt{ggplot2} framework, without the need for objects other than simple data frames.

  \subsection{Data Structure}
  In order for this geometry to work using data frames, there need to be two separate data frames given to the \texttt{geom\_net} function: one for the edge information and one for the vertex information. The vertex data frame should contain all the relevant vertex information. \emph{The only necessary variables are the vertex names/labels and the vertex degree (if undirected) or the vertex in degree and out degree (if directed).  The names and degrees are stored in columns and the rows are the vertices.  Other values of interest, such as grouping variables, for each vertex should be stored as columns in this dataframe, with an observation for each vertex.}
  \par The edge data frame should contain all the relevant edge information. The only necessary variables are the ``from vertex" and ``to vertex" for each edge in the network. The From and To vertices should be listed according to the names of the vertices in the vertex information data frame. Other variables may also be included for each edge, such as the edge weight or grouping variable.  As before, the variables of interest are columns in the data frame and the rows are each edge in the network.%\footnote{I'm not 100\% certain of all of this yet, but so far it seems to me to be the best way to think about networks in terms of data frames.}

>>>>

This network is shown in figure~\ref{fig.cap:madmen_ex}. \emph{Notes: add names, color for gender}
\hh{XXX if you show the code, you should acknowledge it in the write-up. I think, for the motivating example here, it would be OK to leave out the code ... for the examples later on, the code is much more important.}

@@

 In the plot, we can see one central character who has many more relationships than any other character. This vertex represents the main character of the show, Don Draper.
\par There are many kinds of networks, and networks are extensively studied across many disciplines.  Many sociologists study social networks, and many biologists study protein networks. As different as these and the many other disciplines that study networks are, they all need the ability to quickly and effectively visualize networks.  We found the current tools to be lacking in this ability, so we chose to fill this gap by adding network plotting capabilities to the popular and widely used \texttt{R} package \texttt{ggplot2}.
\hh{XXX How widely used is ggplot2 at the moment? We could include some statistics from \url{http://cran-logs.rstudio.com/} to demonstrate this.}

@@

\par The \texttt{GGally} package by \citet{ggally} contains a very useful function written by Fran\c{c}ois Briatte and Moritz Marbach called \texttt{ggnet} that does allow for fairly straightforward customization of these three necessary graph attributes.
Coloring the vertices or edges in a graph is a quick and easy way to visualize grouping and can help with pattern or cluster detection. The vertices in a network and the edges between them compose the structure of a network, and being able to discover patterns among them visually is a key part of network analysis. Viewing multiple layouts of the same network can also help reveal patterns or clusters that would not be discovered when only viewing one layout or analyzing only an adjacency matrix. The \texttt{ggnet} function, however, requires the graph input to be a \texttt{network} object according to the \texttt{network} package.  Our work builds off of the \texttt{ggnet} function and presents an intuitive \texttt{geom} for network visualization within the \texttt{ggplot2} framework, without the need for objects other than simple data frames.

  \subsection{Data Structure}
  \hh{XXX we need a link between the two data frames: we have from_id and to_id in the edge data frame, but we also need an 'id' variable in the vertex data. We don't officially have that at the moment - in the code there is a soft requirement (new as of Mar 20) that one of the columns in the edge data set has to be called 'label' ... that's a bit stoopid, but I couldn't think of a better name. It would make more sense to call this id or edgeid or something along those lines - let's talk about it, and make it a formal requirement. }
  In order for this geometry to work using data frames, there need to be two separate data frames given to the \texttt{geom\_net} function: one for the edge information and one for the vertex information. The vertex data frame should contain all the relevant vertex information. \emph{The only necessary variables are the vertex names/labels and the vertex degree (if undirected) or the vertex in degree and out degree (if directed).  The names and degrees are stored in columns and the rows are the vertices.  Other values of interest, such as grouping variables, for each vertex should be stored as columns in this dataframe, with an observation for each vertex.}
  \par The edge data frame should contain all the relevant edge information. The only necessary variables are the ``from vertex" and ``to vertex" for each edge in the network. The From and To vertices should be listed according to the names of the vertices in the vertex information data frame. Other variables may also be included for each edge, such as the edge weight or grouping variable.  As before, the variables of interest are columns in the data frame and the rows are each edge in the network.%\footnote{I'm not 100\% certain of all of this yet, but so far it seems to me to be the best way to think about networks in terms of data frames.}
